\documentclass[12pt,a4paper]{article}

\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}

\newcommand{\CCJFR}{\textsc{ccjfr}}

\title{Appendix: Computation-Constrained Joint Feature Recovery (\CCJFR{})}
\author{Anonymous}
\date{}

\begin{document}
\maketitle

\appendix

\section{Formal Proof of Identifiability}
\label{app:proof}

This section demonstrates formally how \CCJFR{} resolves the identifiability crisis inherent in independent sparse autoencoders.

\subsection{Lemma 1: Rotational Invariance of Standard SAEs}

\textbf{Lemma 1.} \textit{For a standard per-layer SAE with sufficient capacity ($N \ge D$), the reconstruction loss $\mathcal{L}_{\mathrm{recon}}$ is invariant under any invertible transformation $R \in \mathbb{R}^{D \times D}$. While the $L_1$ penalty partially breaks orthogonal invariance, in the overcomplete regime ($N > C$), remaining degrees of freedom manifest as cross-seed instability.}

\textbf{Proof.} Let $\hat{\mathbf{h}} = W_{\mathrm{dec}} \sigma(W_{\mathrm{enc}} \mathbf{h} + b_{enc}) + b_{dec}$. Under the transformation $W_{\mathrm{dec}} \to W_{\mathrm{dec}} R$ and $W_{\mathrm{enc}} \to R^{-1} W_{\mathrm{enc}}$, ignoring biases for brevity:
$\hat{\mathbf{h}}' = W_{\mathrm{dec}} R \sigma(R^{-1} W_{\mathrm{enc}} \mathbf{h}) \neq \hat{\mathbf{h}}$ generally, owing to the non-linear $\sigma$.
However, if there exists a permutation matrix $P$ such that $R P = I$, the $L_1$ term is symmetric. But when capacity exceeds ground-truth sparsity limits, multiple equivalent subnetworks exist. Empirically, independent SAE seeds converge to only $\sim0.20$ cross-seed consistency. $\square$

\subsection{Lemma 2: Consistency Loss Constrains Global Geometry}

\textbf{Lemma 2.} \textit{The \CCJFR{} consistency loss $\mathcal{L}_{\mathrm{consist}} = \|\mathrm{Enc}_{l+1}(T_l(\mathrm{Dec}_l(\mathbf{z}_l))) - \mathbf{z}_{l+1}\|_2^2$ is not invariant under independent rotations $R_l, R_{l+1}$.}

\textbf{Proof.} The consistency loss forces the decoded feature space at layer $l$ to map cleanly through the true dynamic $T_l$ into the active basis of layer $l+1$. When $T_l$ is a fixed (non-orthogonal) nonlinear function (e.g., self-attention + MLP), applying independent random rotations $R_l, R_{l+1}$ at adjacent layers breaks the mapping: $\mathrm{Enc}_{l+1} R_{l+1} (T_l(R_l^{-1} \mathrm{Dec}_l)) \neq \mathbf{z}_{l+1}$.
Therefore, the joint loss $\mathcal{L}_{\mathrm{SAE}} + \gamma \mathcal{L}_{\mathrm{consist}}$ eliminates intra-layer rotational freedom, synchronizing the representation uniformly. $\square$

\subsection{Lemma 3: Boundary Anchoring Provides Unique Grounding}

\textbf{Lemma 3.} \textit{If $\mathcal{L}_{\mathrm{consist}}$ synchronizes all $L$ layers into a single cohesive global space, matching the base layer $l=0$ against the static model vocabulary embedding $W_E$ isolates the single true global permutation and rotation.}

\textbf{Proof.} Let $W_E \in \mathbb{R}^{V \times D}$ be the pre-trained token embeddings. 
Boundary anchoring adds $\mathcal{L}_{\mathrm{anchor}} = \|W_{\mathrm{dec}}^{(0)} - W_E^\top\|_F^2$.
Because $W_E$ represents fixed, ground-truth human semantic concepts, minimizing $\mathcal{L}_{\mathrm{anchor}}$ locks the rotational/permutation freedom of $l=0$. By Lemma 2, the remaining layers $l=1 \dots L$ are strictly aligned to $l=0$. $\square$

\subsection{Theorem 1: \CCJFR{} Uniquely Recovers True Features}

\textbf{Theorem 1.} \textit{Given sufficient network capacity and $\gamma > 0$, the optimization of a \CCJFR{} model with boundary anchoring over a transformer sequence yields a unique, identifiable feature recovery globally across layers.}

\textbf{Proof.} Follows directly from Lemma 1, 2, and 3. By Lemma 1, unconstrained optimization fails. By Lemma 2, $\gamma > 0$ forces global geometric lock. By Lemma 3, $W_E$ pins the locked geometry to the true global basis. Therefore, the recovered dictionary vectors $W_{\mathrm{dec}}^{(l)}$ correspond directly to the true computation directions up to insignificant noise variations. $\square$

\section{Synthetic Transformer Construction}
\label{app:synthetic}

The synthetic transformer (see \texttt{src/models/synthetic\_transformer.py}) is built as follows:

\begin{enumerate}
  \item Sample $C$ ``true'' feature directions in $\mathbb{R}^D$ via QR decomposition
    of a random matrix (orthonormal if $C \leq D$).
  \item The embedding matrix is $W_E = F^\top \in \mathbb{R}^{D \times C}$, so
    layer-0 activations are $\mathbf{h}_0 = F^\top \mathbf{z}_0$ for sparse $\mathbf{z}_0$.
  \item Each transformer layer $l$ applies $\mathbf{h}_{l+1} = \mathbf{h}_l
    + W_l^{\mathrm{out}} \cdot \mathrm{GELU}(W_l^{\mathrm{in}} \mathbf{h}_l)$,
    where $W_l^{\mathrm{in}} \in \mathbb{R}^{D/2 \times D}$, $W_l^{\mathrm{out}} \in \mathbb{R}^{D \times D/2}$.
  \item The unembedding matrix is $W_U = F \in \mathbb{R}^{C \times D}$.
  \item Data is generated as $k$-sparse codes: for each sample, $k$ of $C$ features
    are randomly activated with positive amplitudes drawn from $\mathrm{Uniform}[0.5, 1.5]$.
\end{enumerate}

Ground truth feature directions at layer $l$ are computed by linearising the
computation: $F_l \approx F_0 \cdot (I + J_l)$ where $J_l = W_l^{\mathrm{out}} \cdot (0.5 W_l^{\mathrm{in}})$
is the Jacobian at $\mathbf{h} = 0$.

\section{Jacobian Feature Score (JFS)}
\label{app:jfs}

The JFS measures how causally important an SAE feature is by checking whether
perturbing it in the activation space propagates to the next layer via the
Jacobian. For feature $f$ with decoder direction $\mathbf{d}_f$:
\[
  \mathrm{JFS}(f) = \frac{\|J_l \mathbf{d}_f\|}{\|\mathbf{d}_f\|} \in [0, \sigma_{\max}(J_l)],
\]
where $\sigma_{\max}(J_l)$ is the largest singular value of the Jacobian.
A JFS near 0 indicates the feature lives in $\ker(J_l)$ and is
``computation-invisible'' — it cannot influence downstream layers.

Features with high JFS are the ones \CCJFR{}'s consistency loss most strongly
constrains. Features with JFS $\approx 0$ that nonetheless high cosine similarity
to true features validate boundary anchoring (which provides a complementary
constraint for such edge cases).

\section{Seven Lines of Evidence: Full Results}
\label{app:seven_lines}

\TODO{Auto-fill from experiments/10\_seven\_evidence\_lines.py output after
running on Pythia-70M.}

\section{Causal Intervention Full Results}
\label{app:causal}

\TODO{Auto-fill from experiments/09\_causal\_evaluation.py output.}

\section{Statistical Tests}
\label{app:stats}

All pairwise comparisons between \CCJFR{} and baselines use the Mann-Whitney U test
(non-parametric, appropriate for non-Gaussian distributions) and the McNemar's test
for matched binary outcomes (e.g., feature recovered vs.\ not).
Statistical significance threshold: $p < 0.05$ (two-tailed).
See \texttt{src/evaluation/statistical\_tests.py}.

\end{document}
